An Algorithmic Introduction to Lagrangian Coherent Structures

### 2014-01-31 main.tex
Remove FTLE figure from double gyre example
The FTLE is moved to the lambda-line convergence figure as a
background.
Minor revisions throughout article

LCSs are used in fluid dynamics to capture essential flow features and thereby simplify the description of complex fluid flows. More specifically, LCSs are particularly well adapted to analyze two dimensional finite time turbulent flows. LCSs identify transport barriers, repelling and attracting structures and vortex boundaries. Application areas for LCSs are oceanic and atmospheric flows, biological applications\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor} and mechanical systems\parencite{hadjighasem13:_detec_kam}.

The finite-time Lyapunov exponent (FTLE) gives one widely-used mathematical definition for LCSs. Although the FTLE is popular, it is possible to find examples where it fails to identify LCSs correctly\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. To remedy these problems, variational calculus principles applied to material curves in a fluid present new approaches to identifying LCSs\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,farazmand13:_attrac_lagran,haller13:_coher_lagran}. This article presents an algorithmic introduction to variational LCSs using a computational engine called LCS Tool\footnote{LCS Tool is free MATLAB software available for download at: \url{https://github.com/jeixav/LCS-Tool}}.

LCS Tool is a library of MATLAB functions that perform the computations to identify LCSs in two dimensional finite time flows. LCS Tool aims to initiate scientists to LCSs. Existing LCS software programs are:
\begin{itemize}
\item \emph{MANGEN}\parencite{lekien03:_time}, calculates stability manifolds adaptively in two dimensional finite time velocity fields. Includes a graphical user interface and uses MPI for parallel calculations.
\item \emph{LCS MATLAB Kit}\parencite{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\parencite{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic and dataset defined velocity definitions.
\item \emph{FlowVC}\parencite{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
\end{itemize}
These packages are based on the FTLE; the objective of LCS Tool and this article is to encourage and facilitate variational LCS analysis.

@@

The fluid flows considered are defined in two dimensions by a velocity field:
\begin{align*}
\dot{\boldsymbol x} = \boldsymbol v(\boldsymbol x,t), && \boldsymbol x \in U \subset \reals^2, && t \in [t_-,t_+].
\end{align*}
Integration yields the flow map, defined between an initial time $t_0$ and a final time $t$ within $[t_-,t_+]$:
\[
F_{t_0}^t(\boldsymbol x_0) := \boldsymbol x(t,t_0,\boldsymbol x_0).
\]
The time interval used to detect LCSs is left as a choice; it can be the complete interval for which the flow is defined, or data is available, or it can be a chosen subset.

The right Cauchy-Green strain tensor measures Lagrangian strain in the velocity field and is defined:
\[
C_{t_0}^t(\boldsymbol x_0) = \left[\nabla F_{t_0}^t(\boldsymbol x_0)\right]^T \nabla F_{t_0}^t(\boldsymbol x_0).
\]
The eigenvalues of the Cauchy-Green strain tensor are denoted by $\lambda_{1,2}$ and the eigenvectors by $\boldsymbol \xi_{1,2}$. The eigenvalues satisfy $0 < \lambda_1 \leq \lambda_2$ and the eigenvectors satisfy $\boldsymbol \xi_1 \perp \boldsymbol \xi_2$.

@@

$\lambda$ is a free parameter close to unity that indicates the uniform stretching of the curve $\gamma$, i.e., each tangent vector $\boldsymbol r'(s)$ with length $l_{t_0}$ stretches by a factor $\lambda$ to length $l_{t}$ under the flow map:
\begin{gather*}
l_{t_0}(s) = \sqrt{\langle \boldsymbol r'(s), \boldsymbol r'(s) \rangle},\\
l_t(s) = \sqrt{\langle \boldsymbol r'(s), C_{t_0}^t(\boldsymbol r(s)) \boldsymbol r'(s) \rangle},\\
l_t(s) = \lambda\, l_{t_0}(s).
\end{gather*}
For $\lambda=1$, $\boldsymbol \eta^{\lambda}_\pm$ points in the direction of maximal Lagrangian shear and we obtain uniformly non-stretching curves over the finite time interval $[t_0, t]$. In summary, the boundaries of coherent Lagrangian vortices are closed curves that uniformly stretch by a factor $\lambda$ close to unity when advected by the flow. We call them lambda-lines. The outermost lambda-line is the physically observable boundary of an eddy, generating the tracer pattern of an isolated rotating fluid region.

Note that $\boldsymbol \eta^{\lambda}_\pm$ follows from the variational principle that the first variation of the averaged Lagrangian strain along the searched closed curve vanishes (see \parencite{haller13:_coher_lagran} for full derivations):
\begin{eqnarray*}
Q(\gamma) &=& \frac{1}{\sigma} \int_0^\sigma \frac{l_t(s)}{l_{t_0}(s)}\text{d}s\\
\delta Q(\gamma) &=& 0.
\end{eqnarray*}
This coherence principle means that the searched curve $\gamma$, the coherent eddy boundary, is a stationary point of the averaged Lagrangian strain functional and therefore nearby fluid elements cannot break away from the curve $\gamma$.

\subsection{Hyperbolic LCS}
Hyperbolic LCS are trajectories of the two eigenvector fields $\boldsymbol \xi_{1,2}$ of the Cauchy-Green strain tensor $C_{t_0}^t$. Repelling hyperbolic LCSs, denoted as strainlines, are obtained as solutions of the differential equation
\[
\boldsymbol r_1' = \boldsymbol \xi_1(\boldsymbol r),
\]
and attracting hyperbolic LCSs, denoted as stretchlines, are obtained as a solution of the differential equation
\[
\boldsymbol r_2' = \boldsymbol \xi_2(\boldsymbol r).
\]
A normal perturbation to the strainline segment $\boldsymbol r_1'$ grows under the flow map by a factor $\lambda_2$, i.e., fluid elements are repelled by that rate in normal direction. Equivalently, a normal perturbation to the stretchline segment $\boldsymbol r_2'$ shrinks with factor $\lambda_1$ and fluid elements are attracted.

Similarly to ellitpic LCS, hyperbolic LCS are also based on a variational principle, in this case for the Lagrangian shear
\[
p(s) = \frac{\langle \boldsymbol r'(s), D(\boldsymbol r(s)) \boldsymbol r'(s)\rangle}{\sqrt{\langle \boldsymbol r'(s), C(\boldsymbol r(s)) \boldsymbol r'(s)\rangle \langle \boldsymbol r'(s), \boldsymbol r'(s)\rangle}}
\]
where
\[
D(\boldsymbol x_0) = \frac12[C(\boldsymbol x_0) \Omega - \Omega C(\boldsymbol x_0)], \quad \Omega = \begin{pmatrix}0&-1\\1&0\end{pmatrix}.
\]
The first variation of the average Lagrangian shear along a hyperbolic LCS vanishes
\[
P(\gamma) = \frac{1}{\sigma} \int_0^\sigma p(s) \text{d}s.
\]
\[
\delta P(\gamma) = 0.
\]
They are one type of shearless LCS. For details of the derivation of hyperbolic LCS, see \cite{farazmand13:_shearless}.

@@

This section presents the numerical methods used to compute LCS and explains them step by step. In other words, the algorithms that constitute LCS Tool are presented.

@@

To obtain Lagrangian coherent structures the Cauchy-Green strain tensor must be calculated first. More specifically, the eigenvalues and eigenvectors of the Cauchy-Green strain tensor are needed. The function associated with this calculation in LCS Tool is \lstinline!eig_cgStrain!. The main steps of this function are enumerated in Table~\ref{t:Cauchy-Green algorithm} and are described in detail in the following part. Table~\ref{t:eig_cgStrain syntax} summarises the function syntax.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\\
\hline
\lstinline!derivative! & function handle for flow velocity equations\\
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\\
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default is $10^{-2}$.\\
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default is \lstinline!true!.\\
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default is \lstinline!false!.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration options\\
\hline \hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

We setup a Cartesian grid according to user-defined resolutions. We can find the optimal resolution by an iterative procedure in which we start with a low resolution and double it until convergence of LCSs is observed visually. If the chosen domain only comprises few expected LCS, e.g., one vortex, a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, for larger domains, a higher resolution must be chosen. Furthermore an auxiliary grid is setup. This is illustrated in Figure~\ref{f:main and auxiliary grids}. The improvements the auxiliary grid yields compared to using only the main grid were reported in \textcite{farazmand12:_comput_lagran}. Experience indicates that optimal results are obtained when the auxiliary grid spacing is between 1\% and 10\% of the main grid spacing.

@@

The function \lstinline!eig_cgStrain! provides the option to calculate the eigenvector from the auxiliary grid, but the eigenvalues from the main grid. We have found that for flows defined analytically the eigenvalues should be calculated from the main grid whereas for the flows defined by datasets, using the auxiliary grid gives better results.

As stated, typically we set the main grid resolution of the Cauchy-Green strain tensor to around 500 grid points. For a square domain, this implies the velocity equations must be integrated, after including 4 auxiliary grid points around every main grid point, at 1.25 million grid points. The ``natural'' way to perform these integration calculations would be to solve a two dimensional first-order ODE system at every grid point and iterate over every grid point within a for-loop. With 1.25 million grid points this calculation can take an excessively long time. To alleviate this problem, we solve the flow equations in a vector form where the equations at all grid points are combined into a single system of equations, even though the system is composed of independent blocks of two dimensional first-order ODE systems. MATLAB's \lstinline!ode45! function is used to perform the integration and the flow integration typically takes five to ten minutes with default error tolerances. The drawback of vector form integration is its memory requirements can become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the intrinsic two dimensional form.

The Cauchy-Green strain tensor can have singularities, i.e., $C_{t_0}^t(\boldsymbol x_0) = I$. Such points are isolated in general\parencite{delmarcelle94}, therefore no special accommodations for are made in computations for singularities of the Cauchy-Green strain tensor.

@@

Flows that are incompressible, meaning $\nabla \cdot \boldsymbol v = 0$, satisfy the relation $\lambda_1(\boldsymbol x_0) \lambda_2(\boldsymbol x_0) = 1, \forall \: \boldsymbol x_0 \: \in \: U$\parencite{arnold78:_mathem}. This gives the possibility of imposing incompressibility when calculating the eigenvalues of the Cauchy-Green strain tensor by first calculating $\lambda_2$ and then setting $\lambda_1(\boldsymbol x_0) = 1/\lambda_2(\boldsymbol x_0)$. This procedure should be applied carefully since by changing $\lambda_1$ but not $\xi_1$, the error in satisfying the eigenvalue equation for the Cauchy-Green strain tensor may increase.

At some grid points it may occur that $\lambda_2 < 1$ due to numerical integration errors of the flow velocity equations. By setting integration tolerances to smaller values the number of grid points with $\lambda_2 < 1$ can be reduced. Nonetheless, some grid points can be so close to singularities of the Cauchy-Green strain tensor that satisfying $\lambda_2 \geq 1$ everywhere could be excessively computationally costly. Therefore the function \lstinline!eig_cgStrain! records the number of points at which incompressibility cannot be imposed and this measure can also help to set appropriate integration tolerances.

@@

Velocity fields defined by datasets require pre-processing prior to performing the integration that yields the flow map. An interpolation function must be written to enable evaluating the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. MATLAB's interpolation functions can be integrated into the velocity function. As with analytic cases, care should be taken to ensure the interpolation embedded in the flow velocity integration function works correctly when performing integration in vector form. In section \ref{sec:oceandataset}, we present an example how LCS are computed for an ocean dataset and details of the interpolation function are given.

@@

Shear LCSs, or coherent Lagrangian vortices, are closed orbits in the Lagrangian shear vector fields $\boldsymbol \eta_{\pm}$ defined in Equation~\eqref{eq:etafields}. Since $\lambda = 1$, the closed curves are non-stretching and area preserving. We find closed orbits by integrating lambda-lines starting from a Poincare section and evaluating the first return map. A lambda-line closing onto itself in the starting point is a shear LCS.
The main steps of the algorithm used to calculate shear LCS are enumerated in Table~\ref{t:Shear LCS algorithm} and described in detail in the following part. The syntax of shear LCS functions in LCS Tool can be seen in Table~\ref{t:Shear LCS functions}.

\begin{table}
\begin{center}
\begin{enumerate}
\item Position Poincare sections in flow domain as initial positions of lambda-lines
\item Integrate lambda-lines tangent to $\boldsymbol \eta_{\pm}$
\item Calculate Poincare return map
\item Find closed orbits in Poincare return map
\item Identify outermost closed orbit
\end{enumerate}
\end{center}
\caption{Algorithm to calculate Shear LCS}
\label{t:Shear LCS algorithm}
\end{table}

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\\
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\\
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\\
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\\
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\\
\hline
\lstinline!domain! & array to define flow domain\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\\
\hline
\lstinline!PSList! & structure for Poincare section end-points, number of lambda-lines launched from Poincare section, and maximum closed lambda-line length\\
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare return map. Default is 5.\\
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare return map. Default is $10^{-2}$.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify lambda-line integration options\\
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\\
\hline \hline
\end{tabular}
\caption{LCS Tool shear LCS functions syntax}
\label{t:Shear LCS functions}
\end{table}

The first step is to set the position of Poincare sections \lstinline!poincareSection.endPosition! that are straight lines in regions where closed lambda-lines are expected. The Poincare section must be oriented in a way that the first endpoint is close to the centre of the expected eddy and the second endpoint is outside the expected eddy region. There is no predetermined procedure for deciding where the Poincare sections should be positioned. Possibilities are an examination of the finite-time Lyapunov exponent field of the flow, or some prior knowledge of where coherent eddies could be expected. Additionally, the number of lambda-lines launched from the Poincare section \lstinline!poincareSection.numPoints! must be defined. A good default value is $100$.

The second step is to integrate the lambda-lines to obtain the Poincare return map. Integration of lambda-lines is performed with the $\eta_+$ and $\eta_-$ vector fields defined on the main grid. These vector fields are derived from eigenvalues and eigenvectors as described in~\textcite{farazmand12:_comput_lagran}. The eigenvector fields have orientation discontinuities, therefore the vector field must be locally reoriented for the integration. The algorithm used to perform variable step-size integration in vector fields with orientation discontinuities is enumerated in Table~\ref{t:variable step integration} and illustrated in Figure~\ref{f:variable step integration}. Linear interpolation is used when interpolating $\boldsymbol \eta_\pm$ within a grid element, since using higher order interpolation would necessitate verifying that there are no orientation discontinuities at more than only the four nearest grid points. Detection of orientation discontinuities is achieved by checking the inner product of the vector fields at adjacent grid points. Then, rotations exceeding 90\degree\, are classified as orientation discontinuities and are corrected before linear interpolation. Thus when setting the Cauchy-Green strain tensor main grid resolution, a histogram of eigenvector field rotations may be produced to ensure that all rotations are below 90\degree\, or close to 180\degree.

>>>>

LCSs are used in fluid dynamics to capture essential flow features and thereby simplify the description of complex fluid flows. More specifically, we analyze two dimensional finite time turbulent flows to detect LCSs. LCSs identify transport barriers, repelling and attracting structures and vortex boundaries. Application areas for LCSs are oceanic and atmospheric flows, biological applications\parencite{wilson09:_lagran_reynol,tallapragada11:_lagran}, aeronautics\parencite{tang10:_accur_lagran_hong_kong_inter_airpor} and mechanical systems\parencite{hadjighasem13:_detec_kam}.

The finite-time Lyapunov exponent (FTLE) gives one widely-used mathematical definition for LCSs. Although the FTLE is widely used, it is possible to find examples where it fails to identify LCSs correctly\parencite{haller11:_lagran_coher_struc,norgard12:_secon_lagran_coher_struc}. To remedy these problems, variational calculus applied to material curves in a fluid provides a rigorous approach for LCS detection\parencite{haller11:_lagran_coher_struc,farazmand12:_comput_lagran,haller12:_geodes_theor_trans_barrier_two_dimen_flows,farazmand13:_attrac_lagran,haller13:_coher_lagran}. This article presents an algorithmic introduction to variational LCSs supported by a computational engine called LCS Tool\footnote{LCS Tool is available at: \url{https://github.com/jeixav/LCS-Tool}.}.

LCS Tool is a library of MATLAB functions that perform computations to identify LCSs in two dimensional finite time flows. LCS Tool aims to help scientists perform variational LCS calculations. Other LCS software has been developed before; the list includes:
\begin{itemize}
\item \emph{MANGEN}\parencite{lekien03:_time}, calculates stability manifolds adaptively in two dimensional finite time velocity fields. Includes a graphical user interface and uses MPI for parallel calculations.
\item \emph{LCS MATLAB Kit}\parencite{dabiri09:_lmk} calculates the FTLE from velocity datasets. Includes a graphical user interface.
\item \emph{Newman}\parencite{toit10:_trans} calculates the FTLE in $N$ dimensions. Assists ridge extraction of FTLE fields. Supports analytic and dataset defined velocity definitions.
\item \emph{FlowVC}\parencite{shadden10:_flowvc} is a general purpose LCS platform for two and three dimensional datasets. Parallel calculations are supported with Open\-MP, CUDA and OpenCL.
\end{itemize}
These packages are based on the FTLE; the focus of LCS Tool and this article is to encourage and facilitate variational LCS analysis.

@@

The fluid flows considered are defined in two dimensions by a velocity field:
\begin{align*}
\dot{\boldsymbol x} = \boldsymbol v(\boldsymbol x,t), && \boldsymbol x \in U \subset \reals^2, && t \in [t_-,t_+].
\end{align*}
Integration yields the flow map, defined between an initial time $t_0$ and a final time $t$ within $[t_-,t_+]$:
\[
F_{t_0}^t(\boldsymbol x_0) := \boldsymbol x(t,t_0,\boldsymbol x_0).
\]
The time interval used to detect LCSs is arbitrary; it can be the complete interval for which the flow is defined, or data is available, or it can be a sub-interval.

The right Cauchy-Green strain tensor measures Lagrangian strain in the velocity field and is defined:
\[
C_{t_0}^t(\boldsymbol x_0) = \left[D F_{t_0}^t(\boldsymbol x_0)\right]^T D F_{t_0}^t(\boldsymbol x_0).
\]
The eigenvalues of the Cauchy-Green strain tensor are denoted by $\lambda_{1,2}$ and the eigenvectors by $\boldsymbol \xi_{1,2}$. The eigenvalues satisfy $0 < \lambda_1 \leq \lambda_2$ and the eigenvectors satisfy $\boldsymbol \xi_1 \perp \boldsymbol \xi_2$.

@@

$\lambda$ is a free parameter close to unity that indicates the uniform stretching of the curve $\gamma$, i.e., each tangent vector $\boldsymbol r'(s)$ with length $l_{t_0}$ stretches by a factor $\lambda$ to length $l_{t}$ under the flow map:
\begin{gather*}
l_{t_0}(s) = \sqrt{\langle \boldsymbol r'(s), \boldsymbol r'(s) \rangle},\\
l_t(s) = \sqrt{\langle \boldsymbol r'(s), C_{t_0}^t(\boldsymbol r(s)) \boldsymbol r'(s) \rangle},\\
l_t(s) = \lambda\, l_{t_0}(s).
\end{gather*}
For $\lambda=1$, $\boldsymbol \eta^{\lambda}_\pm$ points in the direction of maximal Lagrangian shear and we obtain uniformly non-stretching curves over the finite time interval $[t_0, t]$. In summary, the boundaries of coherent Lagrangian vortices are closed curves that uniformly stretch by a factor $\lambda$ when advected by the flow. We refer to these curves as lambda-lines. The outermost lambda-line is the physically observable boundary of an eddy, generating the tracer pattern of an isolated rotating fluid region.

Note that the equation for $\boldsymbol \eta^{\lambda}_\pm$ follows from the variational principle that the first variation of the averaged Lagrangian strain along the sought extremal closed curve vanishes (see \parencite{haller13:_coher_lagran} for full derivations):
\begin{eqnarray*}
Q(\gamma) &=& \frac{1}{\sigma} \int_0^\sigma \frac{l_t(s)}{l_{t_0}(s)}\text{d}s\\
\delta Q(\gamma) &=& 0.
\end{eqnarray*}
This coherence principle means the sought curve $\gamma$, the coherent eddy boundary, is a stationary point of the averaged Lagrangian strain functional and therefore nearby fluid elements cannot separate from $\gamma$.

\subsection{Hyperbolic LCS}
Hyperbolic LCSs are trajectories of the eigenvector fields $\boldsymbol \xi_{1,2}$. Repelling hyperbolic LCSs, denoted strainlines, are obtained as solutions of the differential equation
\begin{equation}
\boldsymbol r_1' = \boldsymbol \xi_1(\boldsymbol r),
\label{eq:strainline}
\end{equation}
and attracting hyperbolic LCSs, denoted stretchlines, are obtained as a solution of the differential equation
\[
\boldsymbol r_2' = \boldsymbol \xi_2(\boldsymbol r).
\]
A normal perturbation to the strainline segment $\boldsymbol r_1'$ grows under the flow map by a factor $\lambda_2$, i.e., fluid elements are repelled at that rate in the normal direction. Equivalently, a normal perturbation to the stretchline segment $\boldsymbol r_2'$ shrinks by a factor $\lambda_1$ and fluid elements are attracted.

Similarly to ellitpic LCS, hyperbolic LCS are also based on a variational principle, in this case for the Lagrangian shear
\[
p(s) = \frac{\langle \boldsymbol r'(s), D(\boldsymbol r(s)) \boldsymbol r'(s)\rangle}{\sqrt{\langle \boldsymbol r'(s), C(\boldsymbol r(s)) \boldsymbol r'(s)\rangle \langle \boldsymbol r'(s), \boldsymbol r'(s)\rangle}}
\]
where
\[
D(\boldsymbol x_0) = \frac12[C(\boldsymbol x_0) \Omega - \Omega C(\boldsymbol x_0)], \quad \Omega = \begin{pmatrix}0&-1\\1&0\end{pmatrix}.
\]
The first variation of the average Lagrangian shear along a hyperbolic LCS vanishes
\[
P(\gamma) = \frac{1}{\sigma} \int_0^\sigma p(s) \text{d}s,
\]
\[
\delta P(\gamma) = 0.
\]
They are one type of shearless LCS. For details of the derivation of hyperbolic LCS, see \cite{farazmand13:_shearless}.

@@

This section presents the numerical methods used to compute LCSs and explains them step-wise. In other words, the algorithms that constitute LCS Tool are presented.

@@

To obtain Lagrangian coherent structures the Cauchy-Green strain tensor must be calculated. More specifically, the eigenvalues and eigenvectors of the Cauchy-Green strain tensor are needed. The function associated with this calculation in LCS Tool is \lstinline!eig_cgStrain!. The main steps of this function are enumerated in Table~\ref{t:Cauchy-Green algorithm} and are described in detail below. Table~\ref{t:eig_cgStrain syntax} summarises the syntax of \lstinline!eig_cgStrain!.

@@

\begin{table}
\begin{center}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![cgEigenvector,cgEigenvalue] = eig_cgStrain(derivative,domain,timespan,resolution)!}\\
\hline
\lstinline!derivative! & function handle for flow velocity equations\\
\hline
\lstinline!domain! & $2 \times 2$ array to define flow domain\\
\hline
\lstinline!timespan! & $1 \times 2$ array to define flow timespan\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!auxGridRelDelta! & optional scalar between 0 and 0.5 to specify auxiliary grid spacing. Default is $10^{-2}$.\\
\hline
\lstinline!eigenvalueFromMainGrid! & optional logical to control whether eigenvalues of Cauchy-Green strain are calculated from main or auxiliary grid. Default is \lstinline!true!.\\
\hline
\lstinline!incompressible! & optional logical to specify if incompressibility is imposed. Default is \lstinline!false!.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify flow map integration parameters\\
\hline \hline
\end{tabular}
\caption{Syntax of the function \lstinline!eig_cgStrain!}
\label{t:eig_cgStrain syntax}
\end{center}
\end{table}

We setup a Cartesian grid according to user-defined resolutions. We can find the optimal resolution by an iterative procedure in which we start with a low resolution and double it until convergence of LCSs is observed visually. If the chosen domain only comprises a few expected LCSs, e.g., one vortex, a resolution of about 500 grid point along the longest axis usually gives good results. Otherwise, for larger domains, a higher resolution must be chosen. Furthermore an auxiliary grid is setup. This is illustrated in Figure~\ref{f:main and auxiliary grids}. The improvements the auxiliary grid yields compared to using only the main grid were reported in \textcite{farazmand12:_comput_lagran}. Experience indicates that optimal results are obtained when the auxiliary grid spacing is 1\% to 10\% of the main grid spacing.

@@

The function \lstinline!eig_cgStrain! provides the option to calculate eigenvectors from the auxiliary grid, but eigenvalues from the main grid. We have found that for flows defined analytically the eigenvalues should be calculated from the main grid whereas for the flows defined by datasets, using the auxiliary grid gives better results.

As stated, typically we set the main grid resolution of the Cauchy-Green strain tensor to around 500 grid points. For a square domain, this implies the velocity equations must be integrated, after including 4 auxiliary grid points around every main grid point, at 1.25 million grid points. The natural way to perform these integration calculations would be to solve a two dimensional first-order ODE system at every grid point and iterate over every grid point in a for-loop. With 1.25 million grid points this calculation can take an excessively long time. To alleviate this problem, we solve the flow equations in a vector form where the equations at all grid points are combined into a single system of equations, even though the system is composed of independent blocks of two dimensional first-order ODE systems. MATLAB's \lstinline!ode45! function is used to perform the integration and the flow integration typically takes five to ten minutes with default error tolerances. The drawback of vector form integration is that memory requirements can become excessive at high resolutions. Furthermore, writing the velocity function in vector form is more error-prone than the simpler two dimensional form.

The Cauchy-Green strain tensor can have singularities, i.e., $C_{t_0}^t(\boldsymbol x_0) = I$. Such points are isolated in general\parencite{delmarcelle94}, therefore no special accommodations are made in computations for singularities of the Cauchy-Green strain tensor.

@@

Flows that are incompressible, meaning $D \cdot \boldsymbol v = 0$, satisfy the relation $\lambda_1(\boldsymbol x_0) \lambda_2(\boldsymbol x_0) = 1, \forall \: \boldsymbol x_0 \: \in \: U$\parencite{arnold78:_mathem}. This gives the possibility of imposing incompressibility when calculating the eigenvalues of the Cauchy-Green strain tensor by first calculating $\lambda_2$ and then setting $\lambda_1(\boldsymbol x_0) = 1/\lambda_2(\boldsymbol x_0)$. This procedure should be applied carefully since by changing $\lambda_1$ but not $\boldsymbol \xi_1$, the error in satisfying the eigenvalue equation for the Cauchy-Green strain tensor may increase.

At some grid points it may occur that $\lambda_2 < 1$ due to numerical integration errors of the flow velocity equations. By setting integration tolerances to smaller values the number of grid points with $\lambda_2 < 1$ can be reduced. Nonetheless, some grid points can be so close to singularities of the Cauchy-Green strain tensor that satisfying $\lambda_2 \geq 1$ everywhere could be excessively computationally costly. Therefore the function \lstinline!eig_cgStrain! records the number of points at which incompressibility cannot be imposed and this measure can guide setting appropriate integration tolerances.

@@

Velocity fields defined by datasets require pre-processing prior to performing the integration that yields the flow map. An interpolation function must be written to enable evaluating the velocity function at arbitrary points in $U$ and at arbitrary times between $t_-$ and $t_+$. MATLAB's interpolation functions can be used in the velocity function. In section \ref{sec:oceandataset} we present an example for LCSs computed for an ocean dataset and details of the interpolation function are given.

@@

Shear LCSs, or coherent Lagrangian vortices, are closed orbits in the Lagrangian shear vector fields $\boldsymbol \eta_{\pm}$ defined in Equation~\eqref{eq:etafields}. We find closed orbits by integrating lambda-lines starting from a Poincare section and evaluating the return map. A lambda-line returning to its starting point is a shear LCS. The main steps of the algorithm used to calculate shear LCS are enumerated in Table~\ref{t:Shear LCS algorithm} and described in detail below. The syntax of shear LCS functions in LCS Tool is in Table~\ref{t:Shear LCS functions}.

\begin{table}
\begin{center}
\begin{enumerate}
\item Position Poincare sections in flow domain to specify initial positions of lambda-lines
\item Integrate lambda-lines tangent to $\boldsymbol \eta_{\pm}$ (see Equation~\eqref{eq:etafields}.)
\item Calculate Poincare return map
\item Find closed orbits positions from Poincare return map
\item Identify outermost closed orbit
\end{enumerate}
\end{center}
\caption{Algorithm to calculate Shear LCS}
\label{t:Shear LCS algorithm}
\end{table}

\begin{table}
\begin{tabular}{|c|p{.7\textwidth}|}
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![shearline.etaPos,shearline.etaNeg] = lambda_line(cgEigenvector,cgEigenvalue,lambda)!}\\
\hline
\lstinline!cgEigenvector! & array of Cauchy-Green strain eigenvectors\\
\hline
\lstinline!cgEigenvalue! & array of Cauchy-Green strain eigenvalues\\
\hline
\lstinline!lambda! & scalar lambda value in Equation~\eqref{eq:eta}\\
\hline \hline
\multicolumn{2}{|p{\textwidth}|}{\lstinline![closedOrbits,orbits] = poincare_closed_orbit_multi(domain,resolution,shearline,PSList)!}\\
\hline
\lstinline!domain! & array to define flow domain\\
\hline
\lstinline!resolution! & $1 \times 2$ array to define Cauchy-Green strain main grid resolution\\
\hline
\lstinline!shearline! & structure of arrays of $\eta_+$ and $\eta_-$ values on main grid\\
\hline
\lstinline!PSList! & structure for Poincare section end-points, number of lambda-lines launched from Poincare section, and maximum closed lambda-line length\\
\hline
\lstinline!nBisection! & optional number of bisection steps to refine zero crossings of Poincare return map. Default is 5.\\
\hline
\lstinline!dThresh! & optional threshold to discard discontinuous zero crossings of Poincare return map. Default is $10^{-2}$.\\
\hline
\lstinline!odeSolverOptions! & optional \lstinline!odeset! structure to specify lambda-line integration parameters\\
\hline
\lstinline!periodicBc! & optional $1 \times 2$ logical array to specify periodic boundary conditions. Default is \lstinline![false,false]!.\\
\hline \hline
\end{tabular}
\caption{LCS Tool shear LCS functions syntax}
\label{t:Shear LCS functions}
\end{table}

The first step is to set the position of Poincare sections as straight lines in regions where closed lambda-lines are expected. The Poincare section must be oriented such that the first endpoint is close to the centre of the expected eddy and the second endpoint is outside the expected eddy. There is no definite procedure for deciding the position of Poincare sections. Possibilities include an examination of the finite-time Lyapunov exponent field of the flow, or some prior knowledge of where coherent eddies are be expected. Additionally, the number of lambda-lines launched from the Poincare section, \lstinline!poincareSection.numPoints!, must be defined. A good default is 100.

The second step is to integrate the lambda-lines to obtain the Poincare return map. Integration of lambda-lines is performed with the $\boldsymbol \eta_+$ and $\boldsymbol \eta_-$ vector fields defined on the main grid. These vector fields are derived from eigenvalues and eigenvectors as described in~\textcite{farazmand12:_comput_lagran}. The eigenvector fields have orientation discontinuities, therefore the vector field must be locally reoriented for the integration. The algorithm used to perform variable step-size integration in vector fields with orientation discontinuities is enumerated in Table~\ref{t:variable step integration} and illustrated in Figure~\ref{f:variable step integration}. Linear interpolation is used when interpolating $\boldsymbol \eta_\pm$ within a grid element, since using higher order interpolation would necessitate verifying that there are no orientation discontinuities at more than only the four nearest grid points. Detection of orientation discontinuities is achieved by checking the inner product of the vector fields at adjacent grid points. Then, rotations exceeding 90\degree\, are classified as orientation discontinuities and are corrected before linear interpolation. Thus when setting the Cauchy-Green strain tensor main grid resolution, a histogram of eigenvector field rotations may be calculated to ensure all rotations are well below 90\degree\, or almost 180\degree.
